---
title: "Declan Bradley - Howard Center"
output: html_document
date: "2025-04-22"
---

A demo notebook for Dow Jones News Fund data interns

During the week’s training, you will work on a project called Investigate Your Community to produce detailed story pitches relevant to their summer internship positions. The goal is to have the students research a topic relevant to your newsroom so they can get to work immediately at the internship. 

Here are some examples from last year’s DJNF data journalism training:

Hannah Rosenberger, at ASI Media/Counselor magazine, produced this analysis on name, image and likeness trends: https://hrosenberger.github.io/ASI-NIL-Project/

Shifra Dayak, an intern at the Connecticut Mirror, produced this analysis of environment spill data: https://shifradayak.github.io/DJNF_Merrill/ct_env_spills.html.

Michael Zhang, an intern at the Houston Landing, explored local election data: https://michaelzhangreporter.github.io/Houston_Campaign_Finance/


Here's what your newsroom suggested for a project
It will likely be something related to trains -- either the data we have from RailState or databases of NTSB investigations, accidents from the Federal Railroad Administration or similar. 


https://github.com/smussenden/declan_railstate

After unzipping, either delete the zip file OR add it to the gitignore for the repo

Declan will need to clone the repo and run the script




This notebook helps you get started

# Load software libraries
```{r}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)

```


# Load your data

2024

```{r}

# 
# YOUR_DATA <- read.csv("../data/XXXXX.csv") |> 
#     clean_names()

# data <- read_excel("../data/2024-12-Revenue.xlsx", skip=2) |> 
#     clean_names()

data <- read.csv("../data/2024 Presidential Post General Md Financial Info 2024 Presidential Post General .csv") |> 
    clean_names()


```

Data details:


# Explore the data types
```{r}

#glimpse(YOUR_DATA)

glimpse(data)

```
# Fix dates
```{r}
data <- data |> 
  mutate(date = lubridate::mdy(contribution_date))
```



# Data exploration: dates

```{r}
data |> 
   count(date) |> 
  arrange(desc(n))
```




# Questions 

Write down questions you want answered:



# Data Details
