---
title: "Declan Bradley - Railstate Pitch"
output: html_document
date: "2025-06-02"
author: "Declan Bradley"
---

**Declan Bradley**, [declanrjb\@proton.me](mailto:declanrjb@proton.me){.email}, (616) 914-9525

# Story Pitch

## Key Idea

The Association of American Railroads requires that trains carrying certain loads of hazardous material, referred to as "key trains," travel no faster than 50 mph. An initial analysis of car and train tables, cross-referenced with snapshots of travel speed at each track sensor, has tentatively identified **9,605 instances of high risk cargoes traveling faster than 50 mph**, in violation of this requirement. (See collected data [here](https://github.com/djnf-data-2025/declan-bradley/blob/main/data/violating-trains.csv).) **#rsw comment: what percentage of these hazmat trains are speeding? follow up on derailment reports and find out how many involved excessive speed#**

## Relevance

The Trump administration has implemented cuts to oversight and enforcement across the federal government, making this story especially impactful in a time of increased risk for accidents in the air, at sea, or on the rails.

## Audience

Decision makers, legislators, and rail industry professionals could use these findings as a basis to step up enforcement of rail safety.

## Potential Interviews

-   A train pilot or technician with experience transporting hazardous materials
-   Officials involved in the NTSB (National Transportation Safety Board)
-   Scientific experts on the specific materials found to be frequently moved at high speeds

## Potential Impact

It should be noted that this analysis is preliminary. However, if these violations are verified, they would be serious incidents that could sway public leaders to step up enforcement of safety regulations.

## Surprising Findings

The sheer number of cars classified as key or hazardous is striking. Even after filtering for cars not marked as empty, the data is still too large to process quickly. Additionally, many trains travel over the recommended limit. I found that certain hazardous materials tend to travel faster than others, which I have as yet no explanation for. Finally, the sheer number of key trains violating the speed limit is remarkable. It is worthy of note that this analysis only encodes the easy identifiers of key trains (identifying number and hazmat class code). Other qualifying characteristics, such as nuclear materials, can only add to the number of key trains once encoded, broadening the scope of the findings.

## Prior Coverage

In February 2023, a train derailment in East Palestine OH spilled hazardous materials including vinyl chloride into the surrounding land. The crash devastated the local community, and an ongoing cleanup effort has so far cost more than \$800 million, according to [NPR reporting](https://www.npr.org/2024/02/05/1228772709/east-palestine-train-derailment-norfolk-southern-lawsuit-epa).

## Visuals

An interactive map can display sensor sites where hazardous trains were detected traveling at high speeds. The data already has lat long coordinates, so producing such a map should not be particularly time consuming. The story can also use images and videos from known crash sites such as the East Palestine derailment, although it will be important to clarify whether those crashes occurred under high speed conditions.

## Further Work

This analysis relies on the assumption that a trainId designates a specific journey and not a specific locomotive. That is to say, if a given trainId is carrying hazardous materials, it is carrying those materials at every sensor observation. If trains instead retain consistent IDs while loading and unloading cars it will be more difficult to guarantee that trains are violating safety codes. A train could, for example, detach its hazardous cars at one station, then accelerate before the next sensor, all without violating safety codes. Additionally, the document on which these definitions are based dates to 2011. A thorough story will need to either base its definitions on more recent documentation or verify that these remain the latest guidelines.

## Estimated Delivery

The data is readily available, but analysis will take time. The most time consuming aspect of the project will be finding sources involved in the rail industry willing to speak on the record about these issues. To be cautious, I would estimate a project ready for review by an editor, if not immediate publication, would take until mid July.

# Analysis Code

Code to open the database and establish its connections:

```{r configure_space, eval=FALSE}
library(tidyverse)

# A generic library for database connections
# https://dbi.r-dbi.org/
library(DBI)

# A specific package for the flavor of database you are connecting.  
# For MYSql, the package is library(RMySQL)
# For Postgres, the package is library(RPostgreSQL)
# I'm using a SQLite database for this example
# https://rsqlite.r-dbi.org/
library(RSQLite)

# For R to SQL translation
library(dbplyr)
library(janitor)
library(googlesheets4)
library(readxl)
library(here)
library(leaflet)
library(googledrive)
source('declan_railstate/shared_scripts/helper_functions.R')

placards_crosswalk <- read_csv(here("declan_railstate/other_data/placards_crosswalk_cfr_census_railstate.csv")) %>%
  filter(in_railstate==TRUE) %>%
  select(placard_type, hazmat_class_railstate, hazmat_name_cfr) 

conn <- create_railstate_db_connection(db_type="merged", merged_geography="all_sensors", sensor_number=NULL)

railstate_tables_and_cols <- get_railstate_tables_and_cols(conn=conn, get_count_of_rows_very_slow = FALSE)

railstate_table_connections <- create_table_connections(conn)

```

Code to analyze the resulting data, isolating key trains (as defined [here](https://www.aar.org/wp-content/uploads/2017/12/AAR-US-Hazmat-Instructions-Rail-BOE.pdf)), and identifying their speeds:

```{r analyze_data, eval=FALSE}

cars_w_hazmat_classes <- railstate_table_connections$tHazmat |>
  filter(car_hazmat_info_str != "[]") %>%
  filter(car_hazmat_info_str != "null") %>%
  filter(car_hazmat_info_str != '[{"placardType": null, "hazmatClass": null}]') %>%
  filter(car_hazmat_info_str != '[{"placardType": "EMPTY", "hazmatClass": null}]') %>%
  filter(placardType != "OTHER") %>%
  filter(placardType != "EMPTY")

cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  left_join(railstate_table_connections$tTrainSightings |>
            select(sightingId, trainId))

cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  left_join(railstate_table_connections$tCars |>
            select(sightingId, carPosition, type, isLoaded),
            by=c('sightingId', 'carPosition'))

# presuming 1 means true
cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  mutate(key_car = case_when(
    (placardType %in% c("UN1005", "UN3318")) && (type == "Tank Car") && (is.na(isLoaded) || (isLoaded == 1)) ~ TRUE,
    TRUE ~ FALSE
  )) |>
  mutate(hazard_shipment = case_when(
    ((placardType %in% c("UN1005", "UN3318")) || (hazmatClass %in% c(2.1, 1.1, 1.2))) && (is.na(isLoaded) || (isLoaded == 1)) ~ TRUE,
    TRUE ~ FALSE
  ))

key_cars_per_train <- cars_w_hazmat_classes |>
  select(trainId, key_car, hazard_shipment) |>
  group_by(trainId) |>
  summarize(num_key_cars = sum(key_car))

hazard_shipments_per_train <- cars_w_hazmat_classes |>
  select(trainId, key_car, hazard_shipment) |>
  group_by(trainId) |>
  summarize(num_hazard_shipments = sum(hazard_shipment))

danger_cars_per_train |>
  write.csv('declan_railstate/danger_cars.csv')

message('succesfully saved danger cars')

key_trains <- danger_cars_per_train |>
  filter(num_key_cars >= 5 || num_hazard_shipments >= 20)

message('identified key trains')

key_train_ids <- key_trains |>
  pull(trainId) |>
  unique()

message('saved key trains to vector')

violators <- railstate_table_connections$tTrainSightings |>
  filter((trainId %in% key_train_ids)) |>
  filter(speedMph >= 50)

message('identified violators')

violators |>
  collect() |>
  write.csv('violating-trains.csv', row.names=FALSE)

message('saved violators')

```
