---
title: "Declan Bradley - Railstate Pitch"
output: html_document
date: "2025-06-02"
author: "Declan Bradley"
---

**Declan Bradley**, [declanrjb\@proton.me](mailto:declanrjb@proton.me){.email}, (616) 914-9525

# Story Pitch

## Key Idea

In 2023, a Northern Southern train derailed in East Palestine, OH, spilling toxic materials including hydrogen chloride and phosgene into the local community. Rail sensor data shows that despite the accident - which has so far cost $800 million in an ongoing cleanup effort - Northfolk Southern has continued to operate trains containing significant loads of hazardous material through East Palestine well in excess of their maximum safe speed.

<div class="flourish-embed flourish-scatter" data-src="visualisation/23592486"><script src="https://public.flourish.studio/resources/embed.js"></script><noscript><img src="https://public.flourish.studio/visualisation/23592486/thumbnail" width="100%" alt="scatter visualization" /></noscript></div>

## Relevance

The Trump administration has implemented cuts to oversight and enforcement across the federal government, making this story especially impactful in a time of increased risk for accidents in the air, at sea, or on the rails.

## Audience

Decision makers, legislators, and rail industry professionals could use these findings as a basis to step up enforcement of rail safety.

## Potential Interviews

-   Residents of East Palestine, OH affected by the crash and the ongoing cleanup in their community
-   Officials involved in the NTSB (National Transportation Safety Board)
-   Scientific experts on the specific materials found to be frequently moved at high speeds

## Potential Impact

It should be noted that this analysis is preliminary. However, if these violations are verified, they would be serious incidents that could sway public leaders to step up enforcement of safety regulations. The continued violation of safe speeds by Norfolk Southern **after** a serious crash may signal troublesome business practice.

## Surprising Findings

I did not anticipate finding a direct connection to the East Palestine crash. This result was arrived at without intent in the programming process. New Waterford appeared in the data as the most common sensor, and I did not realize until mapping it that the two were within 10 miles of one another.

## Prior Coverage

In February 2023, a train derailment in East Palestine, OH, spilled hazardous materials including vinyl chloride into the surrounding land. The crash devastated the local community, and an ongoing cleanup effort has so far cost more than \$800 million, according to [NPR reporting](https://www.npr.org/2024/02/05/1228772709/east-palestine-train-derailment-norfolk-southern-lawsuit-epa).

## Visuals

An interactive map can display sensor sites where hazardous trains were detected traveling at high speeds. The data already has lat long coordinates, so producing such a map should not be particularly time consuming. The story can also use images and videos from the East Palestine crash site.

## Further Work

This analysis only encodes the easy [identifiers of key trains](https://github.com/djnf-data-2025/declan-bradley/blob/main/references/aar-instructions.pdf) (identifying number and hazmat class code). Other qualifying characteristics, such as nuclear materials, can only add to the number of key trains once encoded, broadening the scope of the findings.

A key question will be whether the East Palestine crash occurred at speeds exceeding the recommended limit. The New Waterford sensor logs don't go back far enough chronologically to have observed transits on February 3rd, 2023, so further reporting will be needed to establish this answer. 

This analysis also relies on the assumption that a trainId designates a specific journey and not a specific locomotive. That is to say, if a given trainId is carrying hazardous materials, it is carrying those materials at every sensor observation. If trains instead retain consistent IDs while loading and unloading cars it will be more difficult to guarantee that trains are violating safety codes. A train could, for example, detach its hazardous cars at one station, then accelerate before the next sensor, all without violating safety codes. Additionally, the document on which these definitions are based dates to 2011. A thorough story will need to either base its definitions on more recent documentation or verify that these remain the latest guidelines.

## Estimated Delivery

The data is readily available, but analysis will take time. The most time consuming aspect of the project will be finding sources involved in the rail industry willing to speak on the record about these issues. To be cautious, I would estimate a project ready for review by an editor, if not immediate publication, would take until mid July.

# Explanation of Data and Analysis

The Association of American Railroads requires that trains carrying certain loads of hazardous material, referred to as "key trains," travel no faster than 50 mph. By cross-referencing car and train tables, I identified 21,811 serial numbers marking trains operating in the U.S. that meet the [requirements for key train status](https://github.com/djnf-data-2025/declan-bradley/blob/main/references/aar-instructions.pdf). By cross referencing these train ids with rail sensor logs, I was able to identify 565 unique key trains that had been logged operating at least 5 mph above the speed limit by at least one sensor, accounting for about 2.6% of all key trains operating in the U.S. over the last two years. (An additional 5% of key trains have been logged operating between 50 and 55 mph.)

<div class="flourish-embed flourish-chart" data-src="visualisation/23569369"><script src="https://public.flourish.studio/resources/embed.js"></script><noscript><img src="https://public.flourish.studio/visualisation/23569369/thumbnail" width="100%" alt="chart visualization" /></noscript></div>

I narrowed the dataset to only the 77 key trains traveling 10 mph above the limit or more. Fifty of those trains had been logged in Ohio. I found that they had all been logged by a single rail sensor, which I reverse geocoded to New Waterford, Ohio. In consulting a map, I observed, without anticipating this finding, that New Waterford neighbors East Palestine, OH, where a major derailment occurred in 2023. I checked the coordinates and found that rail sensor #255, which recorded all 50 speeding hazard trains in Ohio, is 7.6 miles from the coordinates of the East Palestine crash site. All 50 of the speeding hazard trains were operated by Norfolk Southern railways, which also operated the freight train that derailed in East Palestine. 

<div class="flourish-embed flourish-table" data-src="visualisation/23591220"><script src="https://public.flourish.studio/resources/embed.js"></script><noscript><img src="https://public.flourish.studio/visualisation/23591220/thumbnail" width="100%" alt="table visualization" /></noscript></div>



# Analysis Code

Code to open the database and establish its connections:

```{r configure_space, eval=FALSE}
library(tidyverse)

# A generic library for database connections
# https://dbi.r-dbi.org/
library(DBI)

# A specific package for the flavor of database you are connecting.  
# For MYSql, the package is library(RMySQL)
# For Postgres, the package is library(RPostgreSQL)
# I'm using a SQLite database for this example
# https://rsqlite.r-dbi.org/
library(RSQLite)

# For R to SQL translation
library(dbplyr)
library(janitor)
library(googlesheets4)
library(readxl)
library(here)
library(leaflet)
library(googledrive)
source('declan_railstate/shared_scripts/helper_functions.R')

placards_crosswalk <- read_csv(here("declan_railstate/other_data/placards_crosswalk_cfr_census_railstate.csv")) %>%
  filter(in_railstate==TRUE) %>%
  select(placard_type, hazmat_class_railstate, hazmat_name_cfr) 

conn <- create_railstate_db_connection(db_type="merged", merged_geography="all_sensors", sensor_number=NULL)

railstate_tables_and_cols <- get_railstate_tables_and_cols(conn=conn, get_count_of_rows_very_slow = FALSE)

railstate_table_connections <- create_table_connections(conn)

```

Code to analyze the resulting data, isolating key trains (as defined [here](https://www.aar.org/wp-content/uploads/2017/12/AAR-US-Hazmat-Instructions-Rail-BOE.pdf)), and identifying their speeds:

```{r analyze_data, eval=FALSE}

cars_w_hazmat_classes <- railstate_table_connections$tHazmat |>
  filter(car_hazmat_info_str != "[]") %>%
  filter(car_hazmat_info_str != "null") %>%
  filter(car_hazmat_info_str != '[{"placardType": null, "hazmatClass": null}]') %>%
  filter(car_hazmat_info_str != '[{"placardType": "EMPTY", "hazmatClass": null}]') %>%
  filter(placardType != "OTHER") %>%
  filter(placardType != "EMPTY")

cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  left_join(railstate_table_connections$tTrainSightings |>
            select(sightingId, trainId))

cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  left_join(railstate_table_connections$tCars |>
            select(sightingId, carPosition, type, isLoaded),
            by=c('sightingId', 'carPosition'))

# presuming 1 means true
cars_w_hazmat_classes <- cars_w_hazmat_classes |>
  mutate(key_car = case_when(
    (placardType %in% c("UN1005", "UN3318")) && (type == "Tank Car") && (is.na(isLoaded) || (isLoaded == 1)) ~ TRUE,
    TRUE ~ FALSE
  )) |>
  mutate(hazard_shipment = case_when(
    ((placardType %in% c("UN1005", "UN3318")) || (hazmatClass %in% c(2.1, 1.1, 1.2))) && (is.na(isLoaded) || (isLoaded == 1)) ~ TRUE,
    TRUE ~ FALSE
  ))

key_cars_per_train <- cars_w_hazmat_classes |>
  select(trainId, key_car, hazard_shipment) |>
  group_by(trainId) |>
  summarize(num_key_cars = sum(key_car))

hazard_shipments_per_train <- cars_w_hazmat_classes |>
  select(trainId, key_car, hazard_shipment) |>
  group_by(trainId) |>
  summarize(num_hazard_shipments = sum(hazard_shipment))

danger_cars_per_train |>
  write.csv('declan_railstate/danger_cars.csv')

message('succesfully saved danger cars')

key_trains <- danger_cars_per_train |>
  filter(num_key_cars >= 5 || num_hazard_shipments >= 20)

message('identified key trains')

key_train_ids <- key_trains |>
  pull(trainId) |>
  unique()

message('saved key trains to vector')

violators <- railstate_table_connections$tTrainSightings |>
  filter((trainId %in% key_train_ids)) |>
  filter(speedMph >= 50)

message('identified violators')

violators |>
  collect() |>
  write.csv('violating-trains.csv', row.names=FALSE)

message('saved violators')

```
